The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-13 paper/paper.tex
rewrite qmid

To produce these estimates for the link, we take a bayesian point of view.  It might seem sensible to start by specifying some prior beliefs about the link $q^*$.  However, the link is in many cases `nonidentifiable';  we describe this problem in detail in Appendix \ref{sec:casestudies}.  One consequence of nonidentifiability is that standard bayesian inferences would have no robustness to our precise choice of prior beliefs.  To avoid these problems, we focus on quantities which can be identified from the data.  In particular,

@@

    \item Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$ denote the set of links which are consistent with $p,h$.  Notice that $q^*\in\Theta(p^*,h^*)$, but there may be other links $q\in\Theta(p^*,h^*)$.  Any such link yields the same distribution on what we can observe, $N^X,N^Y$.  Therefore there is simply no way to use data to distinguish among them.  This is the essence of the nonidentifiability problem mentioned above.  So, instead of trying to estimate this inestimable quantity, we will focus on three other quantities which can be identified from data.  Let

@@

        \item $q_{\mathrm{lo},x,y} \triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x)$
        \item $q_{\mathrm{hi},x,y} \triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x)$ 
        \item $q_{\mathrm{mid}} \triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2$

@@

    Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Morever, even when $q^*$ is nonidentifiable, we can still be assured that $q_{\mathrm{lo}}, q_{\mathrm{hi}}$ are identifiable and $q_{\mathrm{lo},x,y} \leq q^*(y|x) \leq q_{\mathrm{hi},x,y}$.  We can also hope that $q_{\mathrm{mid}}$ will provide a single estimate which strikes a middle ground; in particular, $q_{\mathrm{mid}}$ is the link value which is closest to uniform among all links which are consistent with $p^*,h^*$.  

@@

Following the bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge we would like to incorporate.  First, we have observed the data, $N^X,N^Y$.  Second, we know from the MLM assumption that there exists \emph{some} value $q^*$ which is consistent with $p^*,h^*$.   To make the second idea formal, let $D(h|h')$ denote the Kullback-Leibler divergence and $G(p,h) = \min_{h':\ \Theta(p,h')\neq \emptyset} D(h|h')$.  $G(p,h)=0$ is a numerical way to indicate that the MLM assumption can hold.  After incorporating this knowledge, we find ourselevs interested in the conditional distribution

@@

    \hat q \triangleq \mathbb{E}\left[q_{\mathrm{mid}} | N^X,N^Y,G(p^*,h^*)=0\right]

@@

    \mathbb{P}\left(q_{\mathrm{lo},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% \\
    \mathbb{P}\left(q_{\mathrm{hi},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% 

@@

In practice, we were not able to find a way to compute the credible interval and posterior means exactly.  It seems difficult to even sample from the conditional distribution.  Common approaches to this type of problem involve Markov-Chain Monte Carlo and Variational Inference, but we were unable to make these approaches work in practice.  It seems nontrivial to work with the condition $G(p,h)=0$ that formalizes the MLM assumption.  We instead take a somewhat na\"ive approach.  We start by defining $\tilde P,\tilde H$ to carry the conditional distribution of $p^*,h^*$ given the observed data (but not conditioned on the fact that the MLM assumption holds).  We then simply project $\tilde H$ to the space of parameters which are consistent with the MLM assumption.  That is, we take $\hat H$ to be the solution of the minimization problem $\min_{h'}D(\tilde H|h')$, subject to the constraint that $E(\tilde P,h')=0$.  Optimization details can be found in Appendix \ref{sec:optproblems}.  We hope that the distribution of $\tilde P,\hat H$ will approximate the desired posterior.  We then use samples from this distribution to compute what we need.  We asymptotically expect that $\tilde P,\tilde H$ will nearly satisfy the MLM assumption in any case, so this approximation should not make a large difference.  However, we caution that the exact implications of the approximation are difficult to reason about.   In any case, it is a practical solution to a difficult problem.

@@

We applied the MLM to this data to obtain point estimates $\tilde q(y|x)$ and credible intervals $C_{x y}$.  In Figure \ref{fig:allenboot} we visualize these objects for selected values of $x,y$.  Each parameter has a slightly different story.  For example, for the Standard type $x=$`Vip Rspo4' and the Patch type $y=$`Vip,' we have that $C_{x,y}=[.92,1.0]$.  This supports the idea that the true link satisfies $q^*(y|x)\geq .92$: at least 92\% of the cells classified as type `Vip Rspo4' by the Standard method will be classified as `Vip' by the Patch method.  However, for other types there is more ambiguity.  For the Standard type $x=$`Lamp5 Egln 1' and the Patch types $y=$`Lamp' we have $C_{x y}=[0,1.0]$.  The data we have does not give us a definitive answer as to whether cells with Standard type `Lamp5 Egln 1' are being classified with Patch type `Lamp.'

>>>>

To produce these estimates for the link, we take a bayesian point of view.  It might seem sensible to start by specifying some prior beliefs about the link $q^*$.  However, the link is in many cases `nonidentifiable,' i.e.\ even with an asymptotically infinite number of samples it is not necessarily possible to get a perfect estimate for the link.   We describe this problem in detail in Appendix \ref{sec:casestudies}.  One consequence of nonidentifiability is that standard bayesian inferences would have no robustness to our precise choice of prior beliefs.  To avoid these problems, we focus on quantities which can be identified from the data.  In particular,

@@

    \item Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$ denote the set of links which are consistent with $p,h$.  Notice that $q^*\in\Theta(p^*,h^*)$, but there may be other links $q\in\Theta(p^*,h^*)$.  Any such link yields the same distribution on what we can observe, $N^X,N^Y$.  Therefore there is simply no way to use data to distinguish among them.  This is the essence of the nonidentifiability problem mentioned above.  So, instead of trying to estimate this inestimable quantity, we will focus on three other quantities which can be identified from data.  For each fixed value of $p^*,h^*$, we define

@@

        \item $q_{\mathrm{lo},x,y}(p^*,h^*) \triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x)$
        \item $q_{\mathrm{hi},x,y}(p^*,h^*) \triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x)$ 
        \item $q_{\mathrm{mid}}(p^*,h^*) \triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} \left(q(y|x) - \mathrm{Uniform}(y)\right) ^2$

@@

    where $\mathrm{Uniform}(y)$ indicates the uniform probability mass function for $y$.  Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known.  Indeed, the set $\Theta$ is a convex polytope, so the first two problems are linear programs and the last problem is a strictly convex quadratic program.  Details can be found in Appendix \ref{sec:optproblems}.  Even when $q^*$ is nonidentifiable, we can still be assured that $q_{\mathrm{lo}}, q_{\mathrm{hi}}$ are identifiable and $q_{\mathrm{lo},x,y} \leq q^*(y|x) \leq q_{\mathrm{hi},x,y}$.  We can also hope that $q_{\mathrm{mid}}$ will provide a single estimate which strikes a middle ground; among all links which are consistent with $p^*,h^*$, $q_{\mathrm{mid}}$ is the one which is closest to uniform in a sum-of-squares sense.

@@

Following the bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge we would like to incorporate.  First, we have observed the data, $N^X,N^Y$.  Second, we know from the MLM assumption that there exists \emph{some} value $q^*$ which is consistent with $p^*,h^*$.   To make the second idea formal, let $D(h|h')$ denote the Kullback-Leibler divergence and $G(p,h) = \min_{h':\ \Theta(p,h')\neq \emptyset} D(h|h')$.  $G(p,h)=0$ is a numerical way to indicate that the MLM assumption can hold.  After incorporating this knowledge, we find ourselves interested in the conditional distribution

@@

    \hat q \triangleq \mathbb{E}\left[q_{\mathrm{mid}}(p^*,h^*) | N^X,N^Y,G(p^*,h^*)=0\right]

@@

    \mathbb{P}\left(q_{\mathrm{lo},x,y}(p^*,h^*) \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% \\
    \mathbb{P}\left(q_{\mathrm{hi},x,y}(p^*,h^*) \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% 

@@

In practice, we were not able to find a way to compute the credible interval and posterior means exactly.  However, given samples from the posterior distribution it would be straightforward to get good estimates, using use monte-carlo approximations.  Unfortunately, it seems difficult to obtain samples from the posterior distribution, $\mathbb{P}(p^*,h^*| N^X,N^Y,G(p^*,h^*)=0)$.   Common approaches to this type of problem involve Markov-Chain Monte Carlo and Variational Inference, but we were unable to make these approaches work in practice.  It seems nontrivial to work with the condition $G(p,h)=0$ that formalizes the MLM assumption.  We instead take a somewhat na\"ive approach.  We start by drawing samples according to
\[
P,H ~ \sim \mathbb{P}(p^*,h^* | N^X,N^Y)
\]
This can be achieved exactly, using the the conjugacy between our prior and the Multinomial distribution.  Notice that these samples do not incorporate our knowledge of the MLM assumption, insofar as they are not conditioned on the event $G(p^*,h^*)=0$.  To approximately remedy this, we define $\tilde H$ as the solution of $\min_{h'}D(H|h')$, subject to the constraint that $E(P,h')=0$.  Optimization details can be found in Appendix \ref{sec:optproblems}.  We use the pair $P,\tilde H$ as \emph{approximate} samples for the distribution $\mathbb{P}(p^*,h^*| N^X,N^Y,G(p^*,h^*)=0)$.  We can repeat this process to produce many samples of $(P,\tilde H)$ and use those samples to produce approximate monte-carlo estimates for $\hat q(y|x),C_{x,y}$.  We asymptotically expect that $P,H$ will nearly satisfy the MLM assumption in any case, so this approximation should not make a large difference.  However, we caution that the exact implications of the approximation are difficult to reason about.   In any case, it is a practical solution to a difficult problem.

@@

We applied the MLM to this data to obtain point estimates $\tilde q(y|x)$ and credible intervals $C_{x y}$.  In Figure \ref{fig:allenboot} we visualize these objects for selected values of $x,y$.  Each parameter has a slightly different story.  For example, for the Standard type $x=$`Vip Rspo4' and the Patch type $y=$`Vip,' we have that $C_{x,y}=[.88,1.0]$.  This supports the idea that the true link satisfies $q^*(y|x)\geq .88$: at least 88\% of the cells classified as type `Vip Rspo4' by the Standard method will be classified as `Vip' by the Patch method.  However, for other types there is more ambiguity.  For the Standard type $x=$`Lamp5 Egln 1' and the Patch types $y=$`Lamp' we have $C_{x y}=[0,1.0]$.  The data we have does not give us a definitive answer as to whether cells with Standard type `Lamp5 Egln 1' are being classified with Patch type `Lamp.'
